---
title: "Mapping sampling locations"
output: html_document
---

This document describes the creation of a sampling map for *Adenanthos*  
It relies on the spatial data from https://gadm.org/download_country.html (choose the Geopackage)  
and IBRA v. 7 data from https://www.environment.gov.au/fed/catalog/search/  

Note: to avoid a massive SVG, plots and labels can be combined/added in Inkscape  

Set the working directory to the folder where the input files are (or links to them)
```{r}
setwd("~/projects/Adenanthos/map")
```

Load up the required libraries and designate the input files  
The `locations.tab` file should have a population/sampling label (e.g. `FOR1`), latitude and longitude, tab separated per line (no header)  
Distribution data (a csv file) needs to have at least three columns named `Species`, `Longitude` and `Latitude`
```{r}
suppressMessages(library(sf))
suppressMessages(library(ggplot2))
suppressMessages(library(ggspatial))
suppressMessages(library(extrafont))
mapfile <- "gadm41_AUS.gpkg"
ibra_file <- "IBRA7_subregions.shp"
coord_data <- "locations.tab"
dist_data <- "A_coords.csv"

# Australia state boundaries
wa_file <- "WA_STATE_POLYGON_shp.shp"
qld_file <- "QLD_STATE_POLYGON_shp.shp"
sa_file <- "SA_STATE_POLYGON_shp.shp"
nt_file <- "NT_STATE_POLYGON_shp.shp"
nsw_file <- "NSW_STATE_POLYGON_shp.shp"
vic_file <- "VIC_STATE_POLYGON_shp.shp"
tas_file <- "TAS_STATE_POLYGON_shp.shp"
act_file <- "ACT_STATE_POLYGON_shp.shp"
```

Import fonts  
Note: if running on a system without Arial (or your desired font), you might need to install the
ttf-mscorefonts-installer package
```{r}
suppressMessages(font_import(prompt = FALSE))
loadfonts()
```

Manually choose and set the limits of the desired output map area (x lon, y lat)  
The limits on the map area are then automatically used to roughly set appropriate dimensions of output 
```{r}
xlimits <- c(116.5, 127.5)
ylimits <- c(-35.2, -31.5)
ratio <- abs((xlimits[2] - xlimits[1]) / (ylimits[2] - ylimits[1]))
if (ratio > 1) { 	# wider than tall
	output_width <- 30
	output_height <- 30 / ratio
} else {	# taller than wide
	output_height <- 30
	output_width <- 30 * ratio
}
```

Set the "species" (should match the `Species` column in the distributions file)
```{r}
species <- c("cuneatus",
	"dobsonii",
	"eyrei",
	"forrestii",
	"glabrescens",
	"ileticos",
	"sericeus"
)
```

Create a colour palette for the "species" (same length and order)
```{r}
col_pal <- c("#cc79a7",
	"#56b4e9",
	"#d55e00",
	"#009e73",
	"#0072b2",
	"#f0e442",
	"#e69f00"
)
```

Read in the locality data (`locations.tab`)  
```{r}
coord_df <- read.table(coord_data, header = FALSE)
colnames(coord_df) <- c("Label", "Latitude", "Longitude")
```

Set the associated colours for the plotting locations (populations/samples)  
Locality naming convention matches "species" or categories (can be searched)  
Ensure the order matches the colour palette order
```{r}
groups <- c("CUN",
	"DOB",
	"HERB",
	"FOR",
	"GLA",
	"ILE",
	"SER"
)
# cycle through the groups
indices <- rep(NA, nrow(coord_df))
index <- 1
for (group in groups) {
	indices[grep(group, coord_df$Label)] <- index
	index <- index + 1
}
# assign colour
coord_df$Colour <- col_pal[indices]
```

Load the distribution coordinates and filter missing data  
Subset the data to match the "species", then set corresponding colours
```{r}
dist_df <- read.csv(dist_data)
filter_dist <- dist_df[!is.na(dist_df$Longitude), ]
filter_dist <- filter_dist[!is.na(filter_dist$Latitude), ]
species_df <- filter_dist[filter_dist$Species %in% species, ]
spp <- species_df$Species
# cycle through the species
indices <- rep(NA, nrow(species_df))
index <- 1
for (entry in species) {
	indices[which(species_df$Species == entry)] <- index
	index <- index + 1
}
# assign colour
species_df$Colour <- col_pal[indices]
```

Convert filtered dataframe(s) latitude and longitude to geo coords  
4283 is the EPSG code for the GDA94 CRS (side note: WGS84 = 4326)
```{r}
locations <- st_as_sf(coord_df, coords = c(x = "Longitude", y = "Latitude"),
	crs = 4283, agr = "constant")
```
Similarly for distribution data, if present
```{r}
distributions <- st_as_sf(species_df, coords = c(x = "Longitude", y = "Latitude"),
	crs = 4283, agr = "constant")
```

Load the base map and use the first one: an outline of Australia  
Transform to GDA94
```{r}
mylayers <- st_layers(mapfile)
mylayer <- st_read(mapfile, layer = mylayers$name[1])
mylayer <- st_transform(mylayer, crs = 4283)
```

Load the IBRA (sub)regions shapefile
```{r}
ibra_shp <- st_read(ibra_file)
ibra_shp <- st_transform(ibra_shp, crs = 4283)
```

Load the state boundaries (already GDA94)
```{r}
act_shp <- st_read(act_file)
nsw_shp <- st_read(nsw_file)
nt_shp <- st_read(nt_file)
qld_shp <- st_read(qld_file)
sa_shp <- st_read(sa_file)
tas_shp <- st_read(tas_file)
vic_shp <- st_read(vic_file)
wa_shp <- st_read(wa_file)
```

Set the basemap outline and fill
```{r}
basemap_outline <- mylayer
basemap <- ibra_shp
```

Create an inset for where this area is in Australia  
Note: plot this separately (see next chunk) to place manually in Inkscape by using the full object rather than the grob  
```{r}
bounding_box <- st_as_sfc(st_bbox(c(xmin = xlimits[1], xmax = xlimits[2],
	ymin = ylimits[1], ymax = ylimits[2]), crs = 4283))
myfullinset <- ggplot(data = mylayer[1]) +
	geom_sf(fill = "#efefef", linewidth = 0.8) +
	geom_sf(data = act_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = nsw_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = nt_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = qld_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = sa_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = tas_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = vic_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = wa_shp, fill = NA, color = "#252525", linewidth = 0.8) +
	geom_sf(data = bounding_box, fill = NA, color = "steelblue", linewidth = 1.5) +
	theme_void() +
	coord_sf(xlim = c(113, 154), ylim = c(-43, -9))
```

Plot the inset separately (to be placed manually)
```{r}
png("map_inset.png", units = "cm", width = 10, height = 10, res = 300)
suppressWarnings(print(myfullinset))
invisible(dev.off())
```

Create the base map with ggplot2
```{r}
mymap <- ggplot(data = basemap) +
	geom_sf(colour = "grey", fill = "#efefef") +
	geom_sf(data = basemap_outline, fill = NA) +
	geom_sf(data = distributions, cex = 2,
		colour = distributions$Colour, shape = 19) +
	geom_sf(data = locations, cex = 1,
		fill = "#000000", shape = 19) +
	annotation_scale(location = "br", width_hint = 0.25, text_cex = 1, text_family = "Arial") +
	xlab(NULL) +
	ylab(NULL) +
	theme(panel.grid.major = element_line(color = "#ffffff", linetype = "solid"),
		panel.background = element_rect(fill = "#ffffff"),
		panel.border = element_rect(color = "black", fill = NA),
		axis.ticks = element_line(color = "#000000"),
		axis.text = element_text(family = "Arial", size = 12, color = "#000000")) +
	coord_sf(xlim = xlimits, ylim = ylimits, expand = FALSE)
```

Preview
```{r}
suppressWarnings(print(mymap))
```

For adding labels and insets manually, create an unlabelled output  
```{r}
png("unlabelled_map.png", units = "cm", width = output_width, height = output_height, res = 600)
suppressWarnings(print(mymap))
invisible(dev.off())
```

For checking and creating labels, create a combination map  
```{r}
combo_map <- mymap +
	geom_sf_label(data = locations,
		aes(label = Label, geometry = geometry, colour = Label, size = 5),
		family = "Arial",
		show.legend = FALSE) +
 	scale_colour_manual(values = setNames(locations$Colour, locations$Label))
```

Alternatively, use the ggrepel package to better see the labels  
```{r}
suppressMessages(library(ggrepel))
combo_map <- mymap +
	geom_label_repel(data = locations,
		aes(label = Label, geometry = geometry, colour = Label, size = 5),
		stat = "sf_coordinates",
		max.overlaps = 20,
		min.segment.length = 0,
		label.padding = 0.2,
		point.padding = 0,
		family = "Arial",
		show.legend = FALSE) +
	scale_colour_manual(values = setNames(locations$Colour, locations$Label))
```

Preview the map as it is  
```{r}
suppressWarnings(print(combo_map))
```

Plot a png for simple reference   
```{r}
png("labelled_map.png", width = output_width, height = output_height, units = "cm", res = 600)
suppressWarnings(print(combo_map))
invisible(dev.off())
```
